New Bayes covariant coupling methods to cope with change of priors in one-vs-one classification

### 2025-08-06 theory.tex
Changes for arboreal coupling
Fix changes up to 2.2 per Peter's proof

Knowledge of vector $\boldsymbol{p}^\textrm{multi}(x)$ is equivalent to knowing the functions $p^\textrm{binary}_{ij}(x)$. Indeed under Assumption \ref{ass:1} we have
\begin{align}
p_{ij}^\textrm{binary}(x) = \frac{p_i^\textrm{multi}(x)}{p_i^\textrm{multi}(x) + p_j^\textrm{multi}(x)}. \label{eq:bt1}
\end{align}

Conversely, it is easy to show that  given values of $p_{ij}(x)$ the system of equations \eqref{eq:bt1} has a unique solution. This result (Proposition \ref{prop:binary2multi}) can be viewed as the fundamental theorem of one-vs-one classification.

\begin{prop}
	 \label{prop:binary2multi}
	 Knowledge of binary Bayes classifiers $p^\textrm{binary}_{ij}$ determines the multiclass Bayes classifier 
	 $\boldsymbol{p}^\textrm{multi}$.
\end{prop}
\begin{proof}
	Given a multiclass distribution $\boldsymbol{p} = (p_1, \ldots,p_K)$ on $K$ classes, let $\alpha_{ij}(\boldsymbol{p})$ denote the odds of $C_i$ to $C_j$:
	\begin{align}
		\alpha_{ij}(\boldsymbol{p}) = \frac{p_i}{p_j} \label{eq:alpha}
	\end{align}

	From \eqref{eq:bt1} we can deduce the odds of classes $C_j$ to $C_i$ for the Bayes classifier since
	\begin{align}
	\alpha_{ji}(\boldsymbol{p}^\textrm{multi}) = \frac{p_j^\textrm{multi}(x)}{p_i^\textrm{multi}(x)}
	=  \frac{d_j(x) P_j}{d_i(x) P_i} = \frac{1}{p^\textrm{binary}_{ij}(x)}- 1.
	\end{align}
	Knowing the odds of $C_2$ to $C_1$, $C_3$ to $C_2$ up to $C_{K}$ to $C_{K-1}$ together with the requirement that probabilities sum to one uniquely determines the whole vector $\boldsymbol{p}^\textrm{multi}(x)$.
\end{proof}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

@@

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations. In parallel to Assumption \ref{ass:1} it is customary to make the following assumption on estimates $r_{ij}$.

@@

A method to solve Bradley-Terry equations is called a \emph{coupling method}. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries representing predicted probabilities $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{sample $x$}\arrow[r, "binary~classifiers"] \arrow[dr,dotted,swap, "multi-class~ prediction"] & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

The system of Bradley-Terry equations is usually inconsistent and additional assumptions are needed to construct an estimator. In Sections \ref{sec:des:exact} and \ref{sec:des:inexact} we review various desiderata used to construct and compare coupling methods.

@@

\begin{itemize}
	\item Given  distributions on the classes $\boldsymbol{p}= (p_1, \ldots, p_K)$, we can define
	\begin{align}
		\sigma(\boldsymbol{p})= (p_{\sigma(1)},p_{\sigma(2)}, \ldots, p_{\sigma(K)}).
	\end{align}
	 \item $\sigma$ also acts on the matrix of binary predictions. Namely, the entry of $\sigma(\boldsymbol{R})$ in $i$-th row and $j$-th column is defined to be $r_{\sigma(i), \sigma(j)}$.
	 \end{itemize}
We say that a coupling method $\boldsymbol{v}$ is \emph{symmetric} if 
\begin{align}
		\sigma(\boldsymbol{v}(\boldsymbol{R})) = \boldsymbol{v}(\sigma(\boldsymbol{R}))\quad\textrm{for any permutation $\sigma$}.
\end{align}

>>>>

Knowledge of vector $\boldsymbol{p}^\textrm{multi}(x)$ is equivalent to knowing the functions $p^\textrm{binary}_{ij}(x)$. Indeed, under Assumption \ref{ass:1} we have
\begin{align}
p_{ij}^\textrm{binary}(x) = \frac{p_i^\textrm{multi}(x)}{p_i^\textrm{multi}(x) + p_j^\textrm{multi}(x)}. \label{eq:bt1}
\end{align}

Conversely, it is easy to show that  given values of $p_{ij}^\textrm{binary}(x)$, the system of equations \eqref{eq:bt1} has a unique solution. This result (Proposition \ref{prop:binary2multi}) can be viewed as the fundamental theorem of one-vs-one classification.

\begin{prop}
	 \label{prop:binary2multi}
	 Under Assumption \ref{ass:1}, knowledge of binary Bayes classifiers $p^\textrm{binary}_{ij}$ determines the multiclass Bayes classifier 
	 $\boldsymbol{p}^\textrm{multi}$.
\end{prop}
\begin{proof}
	Given a multiclass distribution $\boldsymbol{p} = (p_1, \ldots,p_K)$ on $K$ classes, let $\alpha_{ij}(\boldsymbol{p})$ denote the odds of classes $C_i$ to $C_j$:
	\begin{align}
		\alpha_{ij}(\boldsymbol{p}) = \frac{p_i}{p_j} \label{eq:alpha}
	\end{align}

	From \eqref{eq:bt1} we can deduce the odds of  $C_j$ to $C_i$ for the Bayes classifier since
	\begin{align}
	\alpha_{ji}(\boldsymbol{p}^\textrm{multi}) = \frac{p_j^\textrm{multi}(x)}{p_i^\textrm{multi}(x)}
	=  \frac{d_j(x) P_j}{d_i(x) P_i} = \frac{1}{p^\textrm{binary}_{ij}(x)}- 1.
	\end{align}
	Knowing the odds of $C_2$ to $C_1$, $C_3$ to $C_2$ up to $C_{K}$ to $C_{K-1}$ together with the requirement that probabilities sum to one uniquely determines the whole vector $\boldsymbol{p}^\textrm{multi}(x)$.
\end{proof}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

@@

Equations forming system \eqref{eq:bt2} are called Bradley--Terry equations. In parallel to Assumption \ref{ass:1} it is customary to make the following assumption on estimates $r_{ij}$.

@@

A method to solve Bradley--Terry equations is called a \emph{coupling method}. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries representing predicted probabilities $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{observation $x$}\arrow[r, "binary~classifiers"] \arrow[dr,dotted,swap, "multi-class~ prediction"] & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

The system of Bradley--Terry equations is usually inconsistent and additional assumptions are needed to construct an estimator. In Sections \ref{sec:des:exact} and \ref{sec:des:inexact} we review various desiderata used to construct and compare coupling methods.

@@

\begin{itemize}
	\item Given  distributions on the classes $\boldsymbol{p}= (p_1, \ldots, p_K)$, we can define
	\begin{align}
		\sigma(\boldsymbol{p})= (p_{\sigma(1)},p_{\sigma(2)}, \ldots, p_{\sigma(K)}).
	\end{align}
	 \item $\sigma$ also acts on the matrix of binary predictions. Namely, the entry of $\sigma(\boldsymbol{R})$ in $i$-th row and $j$-th column is defined to be $r_{\sigma(i), \sigma(j)}$.
	 \end{itemize}
We say that a coupling method $\boldsymbol{v}$ is \emph{symmetric} if 
\begin{align}
		\sigma(\boldsymbol{v}(\boldsymbol{R})) = \boldsymbol{v}(\sigma(\boldsymbol{R}))\quad\textrm{for any permutation $\sigma$}. \label{eq:symmetry}
\end{align}
